\pkg{plotROC}: A Better Tool for Plotting ROC Curves

### 2015-01-29 Paper/sachs2014plotROC.Rmd
editing and readme

```{r setup, echo = FALSE}
library(knitr)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
opts_chunk$set(dev="tikz", comment = NA)
```

@@

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF versus TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is place along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. We favor the use of gridline and reference lines to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types to labels. Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser. A screen shot of an interactive plot is shown in figure \ref{interact}. Hovering over the display shows the cutoff value at the point nearest to the cursor. Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.

@@

We can create an interactive ROC plot using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The key interactive feature of these plots is that the cutoff values nearest to the mouse cursor are displayed when hovering over the figure. Clicking makes the cutoff label stick until the next click. The character string can be copy-pasted into an html document, or better yet, using \pkg{knitr} [@knitr], we can `cat` the results and use the option `results = 'asis'` so that the interactive plot is displayed correctly. For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the webpage https://sachsmc.github.io/plotROC/. Of note is the `prefix` option, which allows the user to assign each plot a unique identifier. This is necessary to prevent conflicts with multiple plots in a single webpage.

```{r int-no, eval = FALSE}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "a")
  )
```

@@

`plotROC` uses the `ggplot2` package to create ggplot objects. Therefore, themes and annotations can be added to `ggroc` objects in the usual `ggplot` way. A `plotROC` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.

>>>>

```{r setup, echo = FALSE}
library(knitr)
library(stringr)
knit_hooks$set(plot = function(x, options) {
  if ('tikz' %in% options$dev && !options$external) {
    hook_plot_tex(x, options)
  } else hook_plot_md(x, options)
})
opts_chunk$set(dev="tikz", comment = NA)
knit_hooks$set(source = function(x, options){
  if (!is.null(options$verbatim) && options$verbatim){
    opts = gsub(",\\s*verbatim\\s*=\\s*TRUE\\s*", "", options$params.src)
    opts = gsub(",\\s*eval\\s*=\\s*FALSE\\s*", "", opts)
    bef = sprintf('\n\n    ```{r %s}\n', opts, "\n")
    stringr::str_c(
      bef, 
      knitr:::indent_block(paste(x, collapse = '\n'), "    "), 
      "\n    ```\n"
    )
  } else {
    stringr::str_c("\n\n```", tolower(options$engine), "\n", 
      paste(x, collapse = '\n'), "\n```\n\n"
    )
  }
})
```

@@

There are several \proglang{R} packages related to ROC curve estimation that contain dedicated plotting functions. The \pkg{ROCR} package [@rocr] plots the FPF _versus_ TPF, as usual, and then takes the interesting approach of encoding the cutoff values as a separate color scale along the ROC curve itself. A legend for the color scale is place along the vertical axis on the right of the plotting region. The \pkg{pROC} package [@pROC] is mainly focused on estimating confidence intervals and regions for restricted ranges of the ROC curve. The plotting methods therein use the base \proglang{R} plotting functions to create nice displays of the curves along with shaded confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2} [@ggplot2] plotting library to create clear, informative ROC plots, with interactive features for use on the web, and sensible defaults for use in print.

@@

The panels of figure \ref{figure1} illustrate the most common styles of ROC curve plots, and the associated design elements. We favor the use of gridlines and a diagonal reference line to facilitate accurate readings off of the axes. Direct labels are preferred over legends because they omit the additional cognitive step of matching line types or colors to labels. Our \pkg{plotROC} package additionally provides plotting of cutoff values, which are displayed interactively with the web-based output option, and direct labels for print use. Exact confidence regions for points on the ROC curve are optionally calculated and displayed. Additionally, we use axis scales are adjusted to be denser near the margins 0 and 1. In medical applications, it is often necessary to have a very low FPR (less than 10%, for instance), therefore the smaller scales are useful for accurately determining values near the margins. The next section details the usage of the \pkg{plotROC} \proglang{R} package and these features.

@@

The same data.frame `rocdata` can be used to generate an interactive plot to view in Rstudio viewer or web browser by passing it to the `plot_interactive_roc` function. Give the function an optional path to an html file as an argument called `file` to save the interactive plot as a complete web page. A screen shot of an interactive plot is shown in figure \ref{interact}. Hovering over the display shows the cutoff value at the point nearest to the cursor. Clicking makes the cutoff label stick until the next click, and if confidence regions are available, clicks will also display those as grey rectangles.

@@

We can create an interactive ROC plot using the `export_interactive_roc` function, which returns a character string containing the necessary \proglang{HTML} and \proglang{JavaScript}. The key interactive feature of these plots is that the cutoff values nearest to the mouse cursor are displayed when hovering over the figure. Clicking makes the cutoff label stick until the next click. The character string can be copy-pasted into an html document, or better yet, incorporated directly into a dynamic document using \pkg{knitr} [@knitr]. In a \pkg{knitr} document, it is necessary to use the `cat` function on the results and use the chunk option `results = 'asis'` so that the interactive plot is displayed correctly. For documents that contain multiple interactive plots, it is necessary to assign each plot a unique name using the `prefix` argument of `export_interactive_roc`. This is necessary to ensure that the \proglang{JavaScript} code manipulates the correct svg elements. For examples of interactive plots and how to incorporate them into `knitr` documents, see the package vignette (`vignette("examples", package = "plotROC")`) or the web page https://sachsmc.github.io/plotROC/.

```{r int-no, results = 'asis', eval = FALSE, verbatim = TRUE}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "a")
  )
```

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be plotted. Therefore, themes and annotations can be added in the usual \pkg{ggplot2} way. A `plot_journal_roc` figure with a new theme, title, axis label, and AUC annotation is shown in figure \ref{annotate}.

### 2015-01-29 Paper/sachs2014plotROC.tex
editing, moved some figures
editing and readme

\title{\pkg{plotROC}: An \proglang{R} Package to Improve the State of ROC Curve
Plots in the Medical Literature}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{ Michael C. Sachs} %% comma-separated
\Plaintitle{plotROC: An R Package to Improve the State of ROC Curve Plots in the
Medical Literature} %% without formatting

\Shorttitle{\pkg{plotROC}: An \proglang{R} Package for ROC Curve Plots} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. We evaluate current
  trends in the design of ROC curve plots with a literature review. Our
  review suggests that ROC curve plots are often ineffective as
  statistical charts and that poor design obscures the relevant
  information the chart is intended to display. We describe our new
  \proglang{R} package that was created to address the shortcomings of
  existing tools. The package has functions to create informative ROC
  curve plots, with sensible defaults, for use in print or as an
  interactive web-based plot. A web application was developed to reach a
  broader audience of scientists who do not use R.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves have
also been used for a long time in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering the
two possible types of errors: false positives, and false negatives. For
a continuous measurement that we denote as \(M\), convention dictates
that a test positive is defined as \(M\) exceeding some fixed threshold
\(c\): \(M > c\). In reference to the binary outcome that we denote as
\(D\), a good outcome of the test is when the test is positive among an
individual who truly has a disease: \(D = 1\). A bad outcome is when the
test is positive among an individual who does not have the disease
\(D = 0\).

Formally, for a fixed cutoff \(c\), the true positive fractions is the
probability of a test positive among the diseased population:

@@

Since the cutoff \(c\) is not usually fixed in advance, we can plot the
TPF against the FPF for all possible values of \(c\). This is exactly
what the ROC curve is, \(FPF(c)\) on the \(x\) axis and \(TPF(c)\) along
the \(y\) axis. A useless test that is not informative at all in regards
to the disease status has \(TPF(c) = FPF(c)\) for all \(c\). The ROC
plot of a useless test is thus the diagonal line. A perfect test that is
completely informative about disease status has \(TPF(c) = 1\) and
\(FPF(c) = 0\) for all \(c\).

@@

The main purpose of visually displaying the ROC curve is to show the
trade-off between the FPF and TPF as the cutoff \(c\) varies. This can
be useful for aiding viewers in choosing an optimal cutoff for decision
making, for comparing a small number of candidate tests, and for
generally illustrating the performance of the test as a classifier. In
practice, once the FPF and TPF are computed for each unique observed
cutoff value, they can be plotted as a simple line chart or scatter plot
using standard statistical plotting tools. This often leads to the
unfortunate design choice of obscuring the critical and useful third
dimension, the range of cutoff values \(c\).

Another key design element is the use of a diagonal guideline for
comparison. The diagonal guideline serves as a baseline for comparison,
allowing observers to roughly estimate the area between the diagonal and
the estimated ROC curve, which serves as a proxy for estimating the
value of the test for classification above a coin-flip. Likewise,
gridlines inside the plotting region and carefully selected axes allow
for accurate assessment of the TPF and FPF at particular points along
the ROC curve. Many medical studies use ROC curves to compare a
multitude of candidate tests to each other and to the null diagonal. In
those cases, curves need to be distinguished by using a legend combined
with different colors or line types, or direct labels inside the
plotting region.

In the medical literature, FPF and TPF are usually referred to in terms
of the jargon terms Sensitivity and Specificity. Sensitivity is
equivalent to the true positive fraction. Specificity is 1 - FPF, the
true negative fraction. Sometimes, the FPF and TPF are incorrectly
referred to as rates, using the abbreviations FPR and TPR. These are
probabilities and their estimates are proportions, therefore we prefer
the use of the term fraction as opposed to rate.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore,
once the appropriate statistics are estimated, existing plotting
functions can be used to create an ROC curve plot. The addition of axis
labels, titles, legends, and so on, indicate a chart as such. In our
literature review, we observed plots with the distinctive
characteristics of the base plotting functions from Microsoft Office,
\proglang{SAS}, \proglang{SPSS}, and the base \proglang{R} plotting
functions.

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF versus TPF, as usual, and then takes the
interesting approach of encoding the cutoff values as a separate color
scale along the ROC curve itself. A legend for the color scale is place
along the vertical axis on the right of the plotting region. The
\pkg{pROC} package \citep{pROC} is mainly focused on estimating
confidence intervals and regions for restricted ranges of the ROC curve.
The plotting methods therein use the base \proglang{R} plotting
functions to create nice displays of the curves along with shaded
confidence regions.

@@

Anyone giving a cursory look at any of the major medical journals is
likely to find at least one ROC curve plot. We sought assess the usage
of ROC curve plots and to evaluate the design choices made in the
current oncology literature by conducting a small literature review. We
searched Pubmed for clinical trials or observational studies in humans
reported in major oncology journals for the past 10 years for the terms
``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
curve''. The search was conducted on October 8, 2014 and returned 54
papers. From those papers, 47 images were extracted and reviewed. The
exact specifications for the Pubmed query are available in the
supplementary materials.

Each image consisted of a single ROC curve plot or a panel of multiple
plots. Each plot was inspected manually for the following design
features: the number of curves displayed, the type of axis labels
(sensitivity/ 1 - specificity or true/false positive fractions),
presence or absence of grid lines, presence or absence of diagonal guide
line, whether any cutpoint were indicated, the type of curve label
(legend or direct label), and presence of other textual annotations such
as the area under the curve. The numerical results of the survey are
summarized in table \ref{table1}.

@@

The panels of figure \ref{figure1} illustrate the most common styles of
ROC curve plots, and the associated design elements. We favor the use of
gridline and reference lines to facilitate accurate readings off of the
axes. Direct labels are preferred over legends because they omit the
additional cognitive step of matching line types to labels. Our
\pkg{plotROC} package additionally provides plotting of cutoff values,
using hover events in interactive use, and direct labels for print use.
Exact confidence regions for points on the ROC curve are optionally
calculated and displayed. Additionally, we use axis scales are adjusted
to be denser near the margins 0 and 1. In medical applications, it is
often necessary to have a very low FPR (less than 10\%, for instance),
therefore the smaller scales are useful for accurately determining
values near the margins. The next section details the usage of the
\pkg{plotROC} \proglang{R} package and these features.

@@

@@

\begin{verbatim}
library(plotROC)
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
\end{verbatim}

Next we use the \texttt{calculate\_roc} function to compute the
empirical ROC curve. The disease status need not be coded as 0/1, but if
it is not, \texttt{plotROC} assumes (with a warning) that the lowest
value in sort order signifies disease-free status. This returns a
dataframe with three columns: the cutoff values, the TPF and the FPF.

@@

The same data.frame \texttt{rocdata} can be used to generate an
interactive plot to view in Rstudio viewer or web browser. A screen shot
of an interactive plot is shown in figure \ref{interact}. Hovering over
the display shows the cutoff value at the point nearest to the cursor.
Clicking makes the cutoff label stick until the next click, and if
confidence regions are available, clicks will also display those as grey
rectangles.

@@

We can create an interactive ROC plot using the
\texttt{export\_interactive\_roc} function, which returns a character
string containing the necessary \proglang{HTML} and
\proglang{JavaScript}. The key interactive feature of these plots is
that the cutoff values nearest to the mouse cursor are displayed when
hovering over the figure. Clicking makes the cutoff label stick until
the next click. The character string can be copy-pasted into an html
document, or better yet, using \pkg{knitr} \citep{knitr}, we can
\texttt{cat} the results and use the option
\texttt{results = \textquotesingle{}asis\textquotesingle{}} so that the
interactive plot is displayed correctly. For examples of interactive
plots and how to incorporate them into \texttt{knitr} documents, see the
package vignette (\texttt{vignette("examples", package = "plotROC")}) or
the webpage https://sachsmc.github.io/plotROC/. Of note is the
\texttt{prefix} option, which allows the user to assign each plot a
unique identifier. This is necessary to prevent conflicts with multiple
plots in a single webpage.

\begin{verbatim}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "a")
  )
\end{verbatim}

@@

\texttt{plotROC} uses the \texttt{ggplot2} package to create ggplot
objects. Therefore, themes and annotations can be added to
\texttt{ggroc} objects in the usual \texttt{ggplot} way. A
\texttt{plotROC} figure with a new theme, title, axis label, and AUC
annotation is shown in figure \ref{annotate}.

\begin{verbatim}
library(ggplot2)
plot_journal_roc(myrocplot, rocdata) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") + 
  ylab("Sensitivity")
\end{verbatim}

>>>>

\title{\pkg{plotROC}: A Better Tool for Plotting ROC Curves}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{ Michael C. Sachs} %% comma-separated
\Plaintitle{plotROC: A Better Tool for Plotting ROC Curves} %% without formatting

\Shorttitle{\pkg{plotROC}: A Better Tool for Plotting ROC Curves} %% a short title (if necessary)
%% an abstract and keywords
\Abstract{
  Plots of the receiver operating characteristic (ROC) curve are
  ubiquitous in medical research. Designed to simultaneously display the
  operating characteristics at every possible value of a continuous
  diagnostic test, ROC curves are used in oncology to evaluate screening,
  diagnostic, prognostic and predictive biomarkers. We reviewed a sample
  of ROC curve plots from the major oncology journals in order to assess
  current trends in usage and design elements. Our review suggests that
  ROC curve plots are often ineffective as statistical charts and that
  poor design obscures the relevant information the chart is intended to
  display. We describe our new \proglang{R} package that was created to
  address the shortcomings of existing tools. The package has functions to
  create informative ROC curve plots, with sensible defaults, for use in
  print or as an interactive web-based plot. A web application was
  developed to reach a broader audience of scientists who do not use R.
}
\Keywords{ROC curves; graphics; interactive; plots}
\Plainkeywords{ROC curves; graphics; interactive; plots} %% without formatting
%% at least one keyword must be supplied

@@

The Receiver Operating Characteristic (ROC) curve is used to assess the
accuracy of a continuous measurement for predicting a binary outcome. In
medicine, ROC curves have a long history of use for evaluating
diagnostic tests in radiology and general diagnostics. ROC curves have
also been used for decades in signal detection theory.

The accuracy of a diagnostic test can be evaluated by considering two
possible types of errors: false positives, and false negatives. For a
continuous measurement that we denote as \(M\), convention dictates that
a test positive is defined as \(M\) exceeding some fixed threshold
\(c\): \(M > c\). In reference to the binary outcome that we denote as
\(D\), a good outcome of the test is when the test is positive for an
individual who truly has a disease: \(D = 1\). A bad outcome is when the
test is positive for an individual who does not have the disease
\(D = 0\).

Formally, for a fixed cutoff \(c\), the true positive fraction is the
probability of a test positive in the diseased population:

@@

Since the cutoff \(c\) is not fixed in advance, we can plot the TPF
against the FPF for all possible values of \(c\). This is exactly what
the ROC curve is, a plot of \(FPF(c)\) on the \(x\) axis and \(TPF(c)\)
along the \(y\) axis as \(c\) varies. A useless test that is not
informative at all in regards to the disease status has
\(TPF(c) = FPF(c)\) for all \(c\). The ROC plot of a useless test is
thus the diagonal line. A perfect test that is completely informative
about disease status has \(TPF(c) = 1\) and \(FPF(c) = 0\) for all
\(c\).

@@

The main purpose of visually displaying the ROC curve is to show the
trade-off between the FPF and TPF as the cutoff \(c\) varies. This can
be useful for aiding viewers in choosing an optimal cutoff for decision
making, for comparing a small number of candidate tests, and for
generally illustrating the performance of the test as a classifier. In
practice, once the FPF and TPF are computed for each unique observed
cutoff value, they can be plotted as a simple line chart or scatter plot
using standard plotting tools. This often leads to the unfortunate
design choice of obscuring the critical and useful third dimension, the
range of cutoff values \(c\).

Another key design element is the use of a diagonal guideline for
comparison. They allow observers to roughly estimate the area between
the diagonal and the estimated ROC curve, which serves as a proxy for
estimating the value of the test for classification above a useless
test. Likewise, gridlines inside the plotting region and carefully
selected axes allow for accurate assessment of the TPF and FPF at
particular points along the ROC curve. Many medical studies use ROC
curves to compare a multitude of candidate tests to each other. In those
cases, curves need to be distinguished by using different colors or line
types combined with a legend, or direct labels inside the plotting
region.

In the medical literature, FPF and TPF are usually referred to in terms
of the jargon sensitivity and specificity. Sensitivity is equivalent to
the true positive fraction. Specificity is 1 - FPF, the true negative
fraction. Sometimes, the FPF and TPF are incorrectly referred to as
rates, using the abbreviations FPR and TPR. These are probabilities and
their estimates are proportions, therefore we prefer the use of the term
fraction as opposed to rate.

@@

The ROC curve plot is, at the most basic level, a line graph. Therefore,
once the appropriate statistics are estimated, existing plotting
functions can be used to create an ROC curve plot. Viewers can identify
ROC plots through context, by observing the shape of the line, and
through the addition of axis labels, titles, legends, and so on. In our
review of the oncology literature, we observed plots with the
distinctive characteristics of the plotting functions from Microsoft
Office, \proglang{SAS}, \proglang{SPSS}, and the base \proglang{R}
plotting functions \citep{arr}.

There are several \proglang{R} packages related to ROC curve estimation
that contain dedicated plotting functions. The \pkg{ROCR} package
\citep{rocr} plots the FPF \emph{versus} TPF, as usual, and then takes
the interesting approach of encoding the cutoff values as a separate
color scale along the ROC curve itself. A legend for the color scale is
place along the vertical axis on the right of the plotting region. The
\pkg{pROC} package \citep{pROC} is mainly focused on estimating
confidence intervals and regions for restricted ranges of the ROC curve.
The plotting methods therein use the base \proglang{R} plotting
functions to create nice displays of the curves along with shaded
confidence regions. My \pkg{plotROC} package uses the \pkg{ggplot2}
\citep{ggplot2} plotting library to create clear, informative ROC plots,
with interactive features for use on the web, and sensible defaults for
use in print.

@@

Anyone giving a cursory look at any of the major medical journals is
likely to find at least one ROC curve plot. We sought to assess the
usage of ROC curve plots and to evaluate the design choices made in the
current oncology literature by conducting a small literature review. We
searched Pubmed for clinical trials or observational studies in humans
reported in major oncology journals for the past 10 years for the terms
``ROC Curve'' OR ``ROC Analysis'' OR ``Receiver operating characteristic
curve''. The search was conducted on October 8, 2014 and returned 54
papers. From those papers, 47 images were extracted and reviewed. The
exact specifications for the Pubmed query are available in the
manuscript source files.

Each image consisted of a single ROC curve plot or a panel of multiple
plots. Each plot was inspected manually for the following design
features: the number of curves displayed, the type of axis labels
(sensitivity/ 1 - specificity or true/false positive fractions),
presence or absence of grid lines, presence or absence of a diagonal
guide line, whether any cutpoints were indicated, the type of curve
label (legend or direct label), and presence of other textual
annotations such as the AUC. The numerical results of the survey are
summarized in table \ref{table1}.

@@

The panels of figure \ref{figure1} illustrate the most common styles of
ROC curve plots, and the associated design elements. We favor the use of
gridlines and a diagonal reference line to facilitate accurate readings
off of the axes. Direct labels are preferred over legends because they
omit the additional cognitive step of matching line types or colors to
labels. Our \pkg{plotROC} package additionally provides plotting of
cutoff values, which are displayed interactively with the web-based
output option, and direct labels for print use. Exact confidence regions
for points on the ROC curve are optionally calculated and displayed.
Additionally, we use axis scales are adjusted to be denser near the
margins 0 and 1. In medical applications, it is often necessary to have
a very low FPR (less than 10\%, for instance), therefore the smaller
scales are useful for accurately determining values near the margins.
The next section details the usage of the \pkg{plotROC} \proglang{R}
package and these features.

@@

\subsection{Installation and loading}\label{installation-and-loading}

\pkg{plotROC} can be installed from the Comprehensive \proglang{R}
Archive Network, or installed from source.

\begin{verbatim}
install.packages("plotROC")
library(plotROC)
\end{verbatim}

@@

\begin{verbatim}
D.ex <- rbinom(100, size = 1, prob = .5)
M.ex <- rnorm(100, mean = D.ex)
\end{verbatim}

Next we use the \texttt{calculate\_roc} function to compute the
empirical ROC curve. The disease status need not be coded as 0/1, but if
it is not, \texttt{calculate\_roc} assumes (with a warning) that the
lowest value in sort order signifies disease-free status. This returns a
dataframe with three columns: the cutoff values, the TPF and the FPF.

@@

The same data.frame \texttt{rocdata} can be used to generate an
interactive plot to view in Rstudio viewer or web browser by passing it
to the \texttt{plot\_interactive\_roc} function. Give the function an
optional path to an html file as an argument called \texttt{file} to
save the interactive plot as a complete web page. A screen shot of an
interactive plot is shown in figure \ref{interact}. Hovering over the
display shows the cutoff value at the point nearest to the cursor.
Clicking makes the cutoff label stick until the next click, and if
confidence regions are available, clicks will also display those as grey
rectangles.

@@

We can create an interactive ROC plot using the
\texttt{export\_interactive\_roc} function, which returns a character
string containing the necessary \proglang{HTML} and
\proglang{JavaScript}. The key interactive feature of these plots is
that the cutoff values nearest to the mouse cursor are displayed when
hovering over the figure. Clicking makes the cutoff label stick until
the next click. The character string can be copy-pasted into an html
document, or better yet, incorporated directly into a dynamic document
using \pkg{knitr} \citep{knitr}. In a \pkg{knitr} document, it is
necessary to use the \texttt{cat} function on the results and use the
chunk option \texttt{results = \textquotesingle{}asis\textquotesingle{}}
so that the interactive plot is displayed correctly. For documents that
contain multiple interactive plots, it is necessary to assign each plot
a unique name using the \texttt{prefix} argument of
\texttt{export\_interactive\_roc}. This is necessary to ensure that the
\proglang{JavaScript} code manipulates the correct svg elements. For
examples of interactive plots and how to incorporate them into
\texttt{knitr} documents, see the package vignette
(\texttt{vignette("examples", package = "plotROC")}) or the web page
https://sachsmc.github.io/plotROC/.

\begin{verbatim}
```{r int-no, results = 'asis'}
cat(
  export_interactive_roc(myrocplot, cutoffs = rocdata$c, 
                         font.size = "12px", prefix = "a")
  )
```
\end{verbatim}

@@

\pkg{plotROC} uses the \pkg{ggplot2} package to create the objects to be
plotted. Therefore, themes and annotations can be added in the usual
\pkg{ggplot2} way. A \texttt{plot\_journal\_roc} figure with a new
theme, title, axis label, and AUC annotation is shown in figure
\ref{annotate}.

\begin{verbatim}
library(ggplot2)
plot_journal_roc(myrocplot, rocdata) + 
  theme_grey() + 
  geom_abline(intercept = 0, slope = 1, color = "white") + 
  ggtitle("Themes and annotations") + 
  annotate("text", x = .75, y = .25, 
           label = "AUC = 0.80") + 
  scale_y_continuous("Sensitivity") + 
  scale_x_continuous("1 - Specificity")
\end{verbatim}
